---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    # use this syntax to add text on several lines
    address: |
      | First line
      | Second line
    email: \email{name@company.com}
    url: http://rstudio.com
  - name: Gerko Vink
    affiliation: 'Utrecht University'
    # To add another line, use \AND at the end of the previous one as above
  - name: Thomas Debray
    # address: |
    #   | Department of Statistics and Mathematics,
    #   | Faculty of Biosciences,
    #   | Universitat Autònoma de Barcelona
    affiliation: |
      | University Medical Centre Utrecht
    # use a different affiliation in adress field (differently formated here)
    affiliation2: Universitat Autònoma de Barcelona
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  # For running headers, if needed
  short:     "\\pkg{mice}: Multilevel"
abstract: >
  Tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness.
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, R]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
```

# Introduction

This template demonstrates some of the basic LaTeX that you need to know to create a JSS article.

## Code formatting

In general, don't use Markdown, but use the more precise LaTeX commands instead:

* \proglang{Java}
* \pkg{plyr}

One exception is inline code, which can be written inside a pair of backticks (i.e., using the Markdown syntax).

If you want to use LaTeX commands in headers, you need to provide a `short-title` attribute. You can also provide a custom identifier if necessary. See the header of Section \ref{r-code} for example.

# \proglang{R} code {short-title="R code" #r-code}

Can be inserted in regular R markdown blocks.

```{r}
x <- 1:10
x
```

## Features specific to \pkg{rticles} {short-title="Features specific to rticles"}

* Adding short titles to section headers is a feature specific to \pkg{rticles} (implemented via a Pandoc Lua filter). This feature is currently not supported by Pandoc and we will update this template if [it is officially supported in the future](https://github.com/jgm/pandoc/issues/4409).
* Using the `\AND` syntax in the `author` field to add authors on a new line. This is a specific to the `rticles::jss_article` format.
