---
  title: "Removed case studies"
author: "Hanne Oberman"
date: "6-4-2022"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
options(prompt = 'R> ', continue = '+ ')

# seed
set.seed(123)

# packages
library(tidyverse)
library(mice)
library(miceadds)
library(lme4)
library(broom.mixed)
# library(metamisc)
# library(GJRM)
# library(pan)

# functions
`%nin%` <- Negate(`%in%`)
round3 <- function(x){format(round(x, 3), nsmall = 3)}
icc <- function(formula, data){multilevel::ICC1(aov(formula, data))}
# icc for imputed data as function in mice
miceadds::source.all(path = "../R")

# plot parameters
plot_col <- mice:::mdc(1:2) %>% setNames(c("observed", "missing"))

# data
popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE, use.value.labels = FALSE) %>% 
  mutate(school = SCHOOL,
         popular = POPULAR,
         sex = SEX,
         texp = TEXP,
         .keep = "none")
popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
```

# How to handle non-random selection (Case study II: HIV)

Data are simulated and included in the `GJRM` package. We will use the following variables:
  
  - `region` Cluster variable,
- `hiv` HIV diagnosis (0=no, 1=yes),
- `age` Age of the patient,
- `marital` Marital status,
- `condom` Condom use during last intercourse,
- `smoke` Smoker (levels; inclusion restriction variable).

The imputation of these date is based on the toy example from [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R). 

<!-- - main model = analysis model = predicting the prevalence of hiv -->
  <!-- - people who have hiv are more reluctant to taking the test, not to reveal their status -->
  <!-- - solution: take variables from analysis model like condom use, marital status, age, etc. that relate to hiv prevalence -->
  <!-- - exclusion restriction variable = instrumental variable (from economics) = e.g. how kind the nurse was = smoke in this dataset = should be correlated to the missingness indicator, but *not* the variable itself -->
  <!-- - include these variables in the selection equation -->
  <!-- - take the missingness indicator of the variable of interest for the selection equation as DV -->
  <!-- - heckman models estimate BOTH selection and mean models simultaneously, because linked through the error term -->
  <!-- - if the errors are not correlated, the models are not linked at all -> not MNAR, but just MAR -->
  <!-- - if the rho is significantly different from zero, the errors are correlated and the models are related so we have MNAR -->
  <!-- - now, we add clustering -->
  <!-- - copulas = estimate 2 equations separately and use another equation to link them -->
  <!-- - for each of the clusters, est the copula separately, so we have all coeffs for each cluster, incl the rho and sigma (the errors) -->
  <!-- - put together all of the estimates through a meta-analysis model to get population distribution of the estimates -->
  <!-- - use the population parameter estimates to draw marginal parameters from the distribution to get a cluster/study parameter -->
  <!-- - with systematical miss we can est the values of the imputed values based on the marginal distribution -->
  <!-- - optional to add shrinkage: the marginal in account as well -->
  <!-- - the hiv data was simulated by the copula function/model developer -->
  <!-- - we could add a systematically missing variable in the data, because these are all sporadically -->
  
  ```{r hiv, echo=FALSE}
# data
data("hiv", package = "GJRM")
hiv <- hiv %>% 
  select(c("region","hiv", "age", "marital", "condom", "smoke")) %>% 
  mutate(region = as.integer(region))
# md pattern
plot_md_pat(hiv)
```

From the missing data pattern we see that we can set `maxit` to 1, since there is only one variable with missingness.

```{r include=FALSE}
# inclusion restriction variable 
mod_main <- glm(hiv ~ ., family = "binomial", data = hiv) %>% 
  broom::tidy(conf.int = TRUE) %>% 
  .[7, c(2,6,7)] %>% 
  round(., 3)
mod_select <- glm(is.na(hiv) ~ ., family = "binomial", data = hiv) %>% 
  broom::tidy(conf.int = TRUE) %>% 
  .[7, c(2,6,7)] %>% 
  round(., 3)
```

The inclusion restriction variable should be a predictor of the the actual value of the variable of interest, but *not* of missingness indicator for the variable of interest. In this case, the data were simulated to adhere to this requirement. Namely, $\beta_{smoke}$ = `r mod_main$estimate`, 95% CI [`r paste0(mod_main$conf.low, ", ", mod_main$conf.high)`] for the analysis model (`formula = hiv ~ .`), and $\beta_{smoke}$ = `r mod_select$estimate`, 95% CI [`r paste0(mod_select$conf.low, ", ", mod_select$conf.high)`] for the selection model (`formula = is.na(hiv) ~ .`). This means the assumptions for the Heckman-type selection model are met.